Retrofitting Effect Handlers onto OCaml

### 2020-11-13 retro-concurrency.tex
tweaks

The same backtrace can also be obtained by using \emph{frame pointers} instead
of DWARF unwind tables. OCaml allows compiling code with frame pointers, but
they are not enabled by default. OCaml stack tends to be deep with small frames
due to the pervasive use of recursive functions, not all of which are
tail-recursive. Hence, the addition of frame pointers can significantly
increase the size of the
stack\footnote{https://github.com/ocaml/ocaml/issues/5721\#issuecomment-472965549}.
Moreover, not using frame pointers saves two instructions in the function
prologue and epilogue, and makes an extra register (|rbp| on x86) available.
Note that the DWARF unwind information is complementary to the information used
by OCaml to walk the stack for GC.

@@

In this section, we refine the effect handler design for the OCaml use case. The
primary motivation for adding effect handlers to OCaml is to support a non-local
control flow mechanism to implement asynchronous I/O in direct style, generators
and enable task parallelism. These use cases do not resume the continuations
more than once. Hence, our continuations are one-shot, and resuming the
continuation more than once raises |Invalid_argument| exception. It is
well-known that one-shot continuations can be implemented
efficiently~\cite{Bruggeman96}.

@@

The implementation follows the operational semantics, but has a few key
representational differences. Unlike the operational semantics, handlers with
just exception patterns (exception handlers) are implemented differently than
effect handlers. As mentioned in \S~\ref{sec:exn_handlers}, exceptions are
pervasive in OCaml and are so cheap that they are used for local control flow.
Hence, we retain the linked exception handler frame implementation of stock
OCaml exception handlers in Multicore OCaml for performance backwards
compatibility. This differs from other research languages with effect
handlers~\cite{Hillerstrom20,Frank,Eff}, which implement exceptions using
effects (by ignoring the continuation argument in the handler).

>>>>

The same backtrace can also be obtained by using \emph{frame pointers} instead
of DWARF unwind tables. OCaml allows compiling code with frame pointers, but
they are not enabled by default. The OCaml stack tends to be deep with small frames
due to the pervasive use of recursive functions, not all of which are
tail-recursive. Hence, the addition of frame pointers can significantly
increase the size of the
stack\footnote{https://github.com/ocaml/ocaml/issues/5721\#issuecomment-472965549}.
Moreover, not using frame pointers saves two instructions in the function
prologue and epilogue, and makes an extra register (|rbp| on x86\_64) available.
Note that the DWARF unwind information is complementary to the information used
by OCaml to walk the stack for GC.

@@

In this section, we refine the effect handler design for the OCaml use case. The
primary motivation for adding effect handlers to OCaml is to support a non-local
control flow mechanism to implement asynchronous I/O in direct style, generators
and enable task parallelism. These use cases do not resume the continuations
more than once. Hence, our continuations are one-shot, and resuming the
continuation more than once raises an |Invalid_argument| exception. It is
well-known that one-shot continuations can be implemented
efficiently~\cite{Bruggeman96}.

@@

The implementation follows the operational semantics, but has a few key
representational differences. Unlike the operational semantics, handlers with
just exception patterns (exception handlers) are implemented differently than
effect handlers. As mentioned in \S\ref{sec:exn_handlers}, exceptions are
pervasive in OCaml and are so cheap that they are used for local control flow.
Hence, we retain the linked exception handler frame implementation of stock
OCaml exception handlers in Multicore OCaml for performance backwards
compatibility. This differs from other research languages with effect
handlers~\cite{Hillerstrom20,Frank,Eff}, which implement exceptions using
effects (by ignoring the continuation argument in the handler).
